Large-scale peer-to-peer network for mobile platforms: Challenges and Experiences\\

### 2019-02-10 paper.tex
Finalized section 1 and 2

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

\begin{itemize}
\item No secondary communication infrastructure such as WiFi
  access points to fall back on
\item Network nodes are mobile, pattern of mobility is not
  predictable
\item New information may arrive at any time
\item Trustworthy information is scarce, misinformation and
  rumours are common place
\item  Small payloads suffice in many cases and information
  stays relevant for a few minutes
\item  Device battery is a scarce resource, power supply
  for recharging may not be available
\item  Devices are owned by citizens, deployment of
  special-purpose devices is cost prohibitive
\end{itemize}

The above needs are well-recognized in the industry with several
ambitious attempts to address them, e.g., Google Loon
project\footnote{https://loon.co} and Facebook
Aquilla\footnote{https://en.wikipedia.org/wiki/Facebook\_Aquila},
though with limited impact given. Leveraging user mobile devices as
peer nodes for a large-scale deployment has been attempted before,
e.g., the Serval project \cite{gardner-stephen-serval-2011}. Serval
mesh enables p2p communication over on-device WiFi radio, but requires
root access to the device via jailbreaking. Although significant
leassons have been learned through these attempts, a mass-scale p2p
network for such applications remains elusive.

A vast majority of the literature has focused on a traditional model
of stateful and reliable networking. Specifically, the nodes maintain
connectivity with peers, routing is optimized with techniques based on
link state or distance vectors
\cite{clausen-olsr-2003,perkins-aodv-2003} focusing on optimizing the
network utilization. Given the application characteristics above, this
paper identifies main practical challenges associated with modern
device platforms and finds novel ways to overcome them.  This leads to
a new paradigm in p2p networking that is connection-less,
delay-tolerant, and zero-routing.  With extensive experiments and
experience of deploying to almost $10$ million users, MNA represents a
way forward for large-scale p2p networks.

In summary, this paper makes the following key contributions:
\begin{itemize}
\item Identification of a class of applications for p2p and their key
  characteristics
\item A deeper investigation of the practical challenges in supporting
  the above class of applications
\item A novel p2p networking implementation using multiple radio
  channels for modern mobile platforms (Android and iOS)
\item Experimental evaluation and deployment statistics
\end{itemize}

@@

Android and iOS. Experimental evaluation and deployment statistics are
presented in Section~\ref{sec:eval}.  A deeper look at the literature
and contrast to MNA is summarized in Section~\ref{sec:related} with
conclusions in Section~\ref{sec:conclude}.

@@

We focus on four main challenges and describe a system implementation
for addressing them. 

@@

\subsection{Background execution} 

@@

Even when a user is not interacting with the weather app, or worse
yet, when the device is not being used at all, the devices must
actively discover peer devices to send and receive potentially

@@

\cite{gardner-stephen-serval-2011}. We overcame these challenges via
innovative techniques, without resorting to hacks that may violate
user security or Appstore guidelines. With our technique, exchange of
information happens without making network connections. As we describe
later, this is achieved by splitting the messages into small enough
chunks so that they can be stuffed into service advertisements
themselves and broadcast over multiple advertisements in a quick
succession.

@@

@@

Since devices may be offline when new weather information
arrives, MNA on each device must remain active at all times to be able
to discover new information as soon as it arrives. However, the MNA
activity must keep the device battery consumption to a minimum. This
is a challenge because advertising and discovery are power-hungry
operations over the radio channels. Our approach here is
two-pronged. Firstly, via extensive experiments on a large number of
devices, we fine-tune the algorithm governing the intervals at which
discovery and advertisements occur. In a nutshell, receiving new
information during a period causes MNA to be more aggressive in
discovery and advertisement. Similarly, lack of new information for a
period makes the device less aggressive. Secondly, we allow ``wake
up'' messages to be broadcast in the network ahead of an anticipated
severe weather event. When devices receive such messages, they
schedule themselves to remain aggressive during the specified window
of time. Outside of this window, the device can afford to have long
sleep cycles and conserve power. With these techniques, our testing
shows less than 1\% battery consumption per hour on most device
models.

@@

Given the heterogeneity of devices and operating system
distributions in the market, it is quite a challenge to test whether
MNA functions as expected on a given device. Further, running test

@@

passes or fails. This is not a challenge in traditional mobile
application development as the software is confined within a single
connected device. We developed test automation tools and processes
such that multiple devices can be controlled from a single test
station and follow prescribed steps to generate, send, and receives
messages to play out a test scenario. Finally, the framework allows
automated analysis of the observations to determine the test
result. This capability was instrumental in uncovering bugs at a fast
pace to meet the aggressive timeline and avoid the cost of
acquisition. Further, the automation framework is general and can be
expressly applied to test other apps in this fashion.

@@

As devices with mesh advertise on a continuous basis, it may
be possible for a malicious attacker to listen for such advertisements
and reverse-engineer the message formats and protocols used. Then, the
attackers may generate fake messages and advertise them, e.g., a fake
tornado alert. Such false messages cannot be distinguished from the
real ones, and MNA will end up propagating them to as many devices as
possible, ``poisoning'' the network. In general, veracity of such
information cannot be independently verified in open decentralized
distributed systems and previous work on peer-to-peer networks do not
address this challenge. In our approach, since the origin of weather
information is the Weather Channel service, digital signatures can be
attached to all weather alerts broadcast from the service. The mobile
application is distributed with the corresponding public key so that
digital signatures from the service can be verified. If a message
fails such a verification, it is discarded and not forwarded any
further.

\section{System Architecture}
\label{sec:architecture}

\section{Experimental Results}
\label{sec:eval}

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\newcommand{\taskprefix}{CH}
\newcommand{\tasksize}{\normalsize}
\newenvironment{taskenv}[1]{\begin{list}{{\tasksize\sc \theenumi.}}{\usecounter{enumi}
      \settowidth{\labelwidth}{{\tasksize\sc \taskprefix#1-99}}
      \setlength{\leftmargin}{\labelwidth}
      %\addtolength{\leftmargin}{2.0\labelsep}
}}{\end{list}}
\newcounter{task}
\setcounter{task}{-1}
\newcommand{\btask}{\begin{taskenv}{\taskprefix}\setcounter{enumi}{\value{task}}\renewcommand{\theenumi}{\taskprefix$_{\arabic{enumi}}$}}
\newcommand{\etask}{\setcounter{task}{\value{enumi}}\renewcommand{\theenumi}{\arabic{enumi}.}\end{taskenv}}

\begin{document}

@@

%
\btask
%
\item\label{c:0} No secondary communication infrastructure such as WiFi
  access points to fall back on
\item\label{c:1} Network nodes are mobile, pattern of mobility is not
  predictable
\item\label{c:2} New information may arrive at any time
\item\label{c:3} Trustworthy information is scarce, misinformation and
  rumours are common place
\item\label{c:4}  Small payloads suffice in many cases and information
  stays relevant for a few minutes
\item\label{c:5}  Device battery is a scarce resource, power supply
  for recharging may not be available
\item\label{c:6}  Devices are owned by citizens, deployment of
  special-purpose devices is cost prohibitive
%
\etask
%

The above needs are well-recognized in the industry with several
ambitious attempts to address them, e.g., Google Loon
project\footnote{https://loon.co} and Facebook
Aquilla\footnote{https://en.wikipedia.org/wiki/Facebook\_Aquila},
though with limited impact. Leveraging user mobile devices as peer
nodes for a large-scale deployment has been another theme in the prior
works, e.g., the Serval project
\cite{gardner-stephen-serval-2011}. Serval mesh enables p2p
communication over on-device WiFi radio, but requires root access to
the device via jailbreaking. Although significant leassons have been
learned through these attempts, a mass-scale p2p network for such
applications remains elusive.

A vast majority of the literature has focused on a traditional model
of stateful and reliable networking. Specifically, the nodes maintain
connectivity with peers, routing is optimized with techniques based on
link state or distance vectors
\cite{clausen-olsr-2003,perkins-aodv-2003} focusing on optimizing the
network utilization. Given the application characteristics above, this
paper identifies main practical challenges associated with modern
device platforms and finds novel ways to overcome them.  This leads to
MNA --- a new paradigm in p2p networking that employs connection-less,
delay-tolerant, and zero-routing overhead protocols.  MNA is
implemented on both Android and iOS as an SDK and integrated with the
Weather Channel mobile apps. With extensive experiments and experience
of deploying to almost $10$ million users, MNA represents a way
forward for large-scale p2p networks.

In summary, this paper makes the following key contributions:
\begin{itemize}
\item Identification of a class of applications for p2p and their key
  characteristics
\item A deeper investigation of the practical challenges in supporting
  the above class of applications
\item A novel p2p networking SDK using multiple radio
  channels for modern mobile platforms (Android and iOS)
\item Experimental evaluation and deployment statistics
\end{itemize}

@@

Android and iOS in addressing the challenges. Experimental evaluation
and deployment statistics are presented in Section~\ref{sec:eval}.  A
deeper look at the literature and contrast to MNA is summarized in
Section~\ref{sec:related} with conclusions in
Section~\ref{sec:conclude}.

@@

We present the main challenges for modern mobile device platforms in
meeting the classes of applications described above. These have been
uncovered via extensive experiments and in some cases include direct
feedback from the makers of Android and iOS.

@@

\subsection{Operating system constraints}
\label{ch:os}

@@

Due to \ref{c:2}, even when a user is not interacting with the weather
app, or worse yet, when the device is not being used at all, the
devices must discover peer devices to receive and formward potentially

@@

\cite{gardner-stephen-serval-2011}. Clearly, such an approach does not
scale to mass adoption.

@@

\label{ch:power}

@@

Since devices may be offline when new weather information arrives, MNA
processes on each peer must remain active at all times to be able to
discover new information as soon as it arrives. Further, as there is
no back up infrastructure (\ref{c:0}) and a set of peers in range can
change at any time (\ref{c:1}), each peer is responsible for
constantly forwarding available information to other peers via
advertisements. However, due to \ref{c:5}, the MNA activity must keep
the device battery consumption to a minimum. This is a challenge
because advertising and discovery are power-hungry operations over the
radio channels.

@@

Given the heterogeneity of devices and operating system distributions
owned by users (\ref{c:6}), it is challenging to test whether or not
MNA functions as expected on a single device. Further, running test

@@

passes or fails. Further, since range and mobility affect p2p
communications and they are unpredictable (\ref{c:1}), it is important
to run test cases under various mobility patterns across all nodes in
the network. Emulated mobility frameworks such as CORE
\cite{arenholz-core-2008} and EMANE \cite{emane} fall short as the
connection latencies and wireless transmission are specific to device
model and radio. This is not a challenge in traditional mobile
application development as the application functionality is confined
within a single connected device.

@@

As user devices with MNA advertise on a continuous basis over
well-known radio protocols, it may be possible for a malicious
attacker to listen for such advertisements and reverse-engineer the
message formats and protocols used. Then, the attackers may generate
fake messages and advertise them, e.g., a fake tornado alert. Given a
lack of trusted information in such scenarios (\ref{c:3}),
misinformation campaigns can have disastrous consequences.  In open
decentralized systems, such false messages cannot be distinguished
from the real ones, and MNA will end up propagating them to as many
devices as possible, ``poisoning'' the network. In general, veracity
of such information cannot be independently verified in open
decentralized distributed systems and previous work on peer-to-peer
networks do not address this challenge.
\section{System Architecture}
\label{sec:architecture}

We overcame these challenges via innovative techniques, without
resorting to hacks that may violate user security or Appstore
guidelines. With our technique, exchange of information happens
without making network connections. As we describe later, this is
achieved by splitting the messages into small enough chunks so that
they can be stuffed into service advertisements themselves and
broadcast over multiple advertisements in a quick succession.

Our approach here is two-pronged. Firstly, via extensive experiments
on a large number of devices, we fine-tune the algorithms governing
the intervals at which discovery and advertisements occur. In a
nutshell, receiving new information during a period causes MNA to be
more aggressive in discovery and advertisement. Similarly, lack of new
information for a period makes the device less aggressive. Secondly,
we allow ``wake up'' messages to be broadcast in the network ahead of
an anticipated severe weather event. When devices receive such
messages, they schedule themselves to remain aggressive during the
specified window of time. Outside of this window, the device can
afford to have long sleep cycles and conserve power. With these
techniques, our testing shows less than 1\% battery consumption per
hour on most device models.

We developed test automation tools and processes such that multiple
devices can be controlled from a single test station and follow
prescribed steps to generate, send, and receives messages to play out
a test scenario. Finally, the framework allows automated analysis of
the observations to determine the test result. This capability was
instrumental in uncovering bugs at a fast pace to meet the aggressive
timeline and avoid the cost of acquisition. Further, the automation
framework is general and can be expressly applied to test other apps
in this fashion.

In our approach, since the origin of weather information is the
Weather Channel service, digital signatures can be attached to all
weather alerts broadcast from the service. The mobile application is
distributed with the corresponding public key so that digital
signatures from the service can be verified. If a message fails such a
verification, it is discarded and not forwarded any further.
\section{Experimental Results}
\label{sec:eval}
